\LARGE \bf Human-Robot Interaction: Tackling the AI Challenges

### 2014-05-05 paper.tex
Start to rework discussion
Spread out some parts to other places in the paper,
Merged sections as suggested by reviewer 1, cf #15

    \caption{The upper part of the ORO common-sense TBox. All these concepts
    belong to the {\sc OpenCyc} namespace.}

@@

\paragraph{The OpenRobots Ontology} The ORO common-sense ontology has been
designed from two requirements: covering our experimental needs and conforming
as much as possible to the {\sc OpenCyc} upper ontology.

@@

Figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
$(\text{TemporalThing} \sqcap \text{SpatialThing})^{\mathcal{I}} = \emptyset$, with
$\mathcal{I}$ the \emph{interpretation} of our model).

@@

The tree in Figure~\ref{fig|upper_tbox}\footnote{This subset of the ontology is
indeed a tree. This has however not to be the case in general, and, as a matter
of fact, the TBox of the whole ORO common-sense ontology does not form a tree}
is not equally developed in each directions at lower levels. The descendants of
\concept{PartiallyTangibleThing} (\ie what we commonly call \emph{objects}), for
instance, are pictured in Figure~\ref{fig|tangible_things_tbox}. This excerpt
from the ontology makes the bottom-up design process clear: only few types of
\emph{partially tangible} things appear, and those are only subclasses relevant
to the context of service robotics in an human-like environment. For performance
reasons as well as clarity, we do not attempt to proactively extend the
conceptual coverage of the field. We instead opportunistically extend the
common-sense knowledge when required by experiments.

@@

Lastly, the ORO common-sense ontology contains several rules and class
expressions that encode non-trivial inferences. The definition of
\concept{Bottle} as found in the ORO ontology is a typical example:

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Figures~\ref{fig|spark-oro} and~\ref{fig:sparkSubfigures}),
also taking into account the different perspective that each agent has on the
environment.

@@

@@

SPARK is designed to run and continously update the knowledge base (in our
experiments, it runs at about 10Hz). At each steps, it re-computes spatial
relationships and affordances for the whole scene, and send the delta (new
relations and relations that do not hold anymore) to the knowledge base. This
approach may raise scalability concerns (we did not however hit such issues in
our constrainted scenarios), but greatly simplifies the management of the
\emph{dynamics} of the knowledge (\emph{When do I discard outdated knowledge?
When do I update it?}). Since it is equivalent to a reset of the reasoner
domain, it also hides issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).

@@

\paragraph{Symbolic facts production}

@@

\paragraph{Hypotheses on objects states and positions}

@@

\paragraph{Primitive action recognition}

@@

\subsubsection{Ego-centric and allo-centric frames}

@@

@@

While parts of the architecture (SPARK, ORO) have been deployed with external
execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
developed dedicated robot controllers which integrate the whole stack introduced
in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
written in the PRS~\cite{Ingrand1996} language. We also use the Python-based
{\sc pyRobots} collection of tools that provide a large set of high-level
actions and an event-based architecture well suited for prototyping.

They both rely on extensive integration with the knowledge base that serves as the
primary source of semantics for decision making.

@@

The second one involves the {\sc Shary} execution controller and the HATP symbolic
task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
robot try to cooperatively remove objects from a table. behaviours and motions
are fully planned.

@@

Perception of objects is done through a tag-based system and humans are
detected through motion capture. The robot knowledge base is pre-loaded with
the \emph{ORO Commonsense Ontology}.  We next describe two
situations where we can follow the internal robot's reasoning and the
interaction with the users.

@@

This experiment involved a more complex decisional layer where the ORO server
was used in conjunction with the HATP symbolic task planner and the {\sc Shary}
execution controller (the complete software architecture we deployed for this
experiment is the one pictured in Figure~\ref{fig|archi},
page~\pageref{fig|archi}).

@@

\section{Discussion: when Artificial Intelligence enables Human Robot
Interaction}
\label{sect|conclusion}

This last section goes over again the main contributions we have presented
so far: why they are relevant to robotics and artificial intelligence, how do
they compare to similar approaches, how do we support our technical choices and
what remains to be developed/explored in the future.

@@

\subsection{Foundations of cognition for the robot}

Because we build autonomous \emph{companion} robots, designed to live and
interact in environments primarily conceived for humans, we endow our machines
with a set of fundamental cognitive capabilities. The robots perform typically
in dynamics, partially unknown environments, with a variety of places,
artifacts and agents that may carry complex semantics.

While our main focus is on the cognitive abilities required to interact with
humans, \emph{situation assessment}, \emph{spatial reasoning}, \emph{knowledge
representation and reasoning}, \emph{task planning} or \emph{high-level
control} are mandatory cognitive foundations to build upon.

\paragraph{Situation assessment, spatial reasoning}

Real service robots need to ground their decision-making processes in their
perceptions, be it physical (\eg what a camera sees) or not (\eg remote
knowledge bases, dialogue). We call \emph{physical situation assessment} the
cognitive skill that a robot exhibits when it assesses the nature and content of its
surroundings and efficiently monitors its evolution.

Our approach, implemented in the {\sc Spark} module, is a grounded amodal
representation of the robot environment, where different sensing modalities
(vision-based fiducial markers tracking, RGBD human tracking, motion capture)
are merged and where amodal spatial reasoning takes place (for instance,
explanation generation for unexpected perceptions like a disappearing object).

Spatial reasoning~\cite{O'Keefe1999} is a field in its own right, and has been
used for natural language processing for applications such as direction
recognition ~\cite{Kollar2010,Matuszek2010} or language
grounding~\cite{Tellex2010}.~\cite{Skubic2004} present a spatial reasoner
integrated in a robot which computes symbolic positions of objects.

One of the main limit of our current approach is the way semantics are
conveyed: because we use 2D markers for artifacts, we skip most of the object
extraction and recognition challenges. Each object has a unique identifier,
which enable us to load a proper CAD model suitable for 3D spatial reasoning
and prevent recognition mistakes in the knowledge base. While {\sc Spark}
algorithms themselves are ignorant of the input sources, and would work equally
well with a full object recognition stack, we have not tackled it yet.

\paragraph{Robot control}

Cognitive control architectures for robotics is a research field in its own right.
Our originality of our contribution relates to the interleaving of a
semantic-aware architecture designed for autonomy with human-aware decisional
components. To our knowledge, no other robotic architecture covers in depth
those two perspectives.

The Ke Jia project~\cite{Chen2010} is maybe one of the closest approach. It
integrates on a mobile platform a knowledge representation language with
natural language processing, task planning and motion planning. Default and
non-monotonic reasoning has been especially researched within the Ke Jia
project for symbolic task planning~\cite{Ji2011} and underspecified natural
language processing.

The Ke Jia robot has been demonstrated in several tasks involving human-robot
interaction with natural language. These tasks include a task with multiple
\emph{pick \& carry} that are globally optimised, naive physics reasoning via
taught rules or more complex scenarios with the robot delivering drinks, taking
into account changing and mutually exclusive preferences of users.

Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
called \textsc{CRAM} (Cognitive Robot Abstract Machine) that integrates
\textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework based on
Prolog. Its underlying storage is based on an OWL ontology, derived from
\textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated on
several real-world scenarios, where natural language recipes extracted from the
Internet had to be translated into plans and executed in a kitchen environment,
perceived and rebuilt on-line by the robots. This architecture has however not
been deployed in scenarios involving human-robot interactions.

\subsection{Knowledge framework}
\label{krs-discussion}

Knowledge flows in our architecture through a \emph{semantic blackboard}, the
{\sc Oro}-server. We rely on Description Logics (OWL) to represent and
manipulate facts.

Tools to manipulate and reason over ontologies are readily available, mature,
and already well accepted in the robotic community~\cite{Tenorth2009a,
Lim2011}. Alternatives like \emph{Answer Set Programming} have also been
successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
to deal with non-monotonic reasoning. However we did not actually hit any brick wall
while working with OWL ontologies. We may reconsider this choice at a later
stage, but until now it has proven an effective framework to quickly explore
implementations of new cognitive abilities (for instance, it has been
conceptually and technically easy to add support for independent knowledge
models, one per agent the robot interacts with --see section~\ref{sect|tom}).

\subsection{Interacting with humans}

\paragraph{Building a model of a human during the interaction} Perspective
Taking is a human ability which allows one to put him/herself in another
person's point of view. Studied in psychology
literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
interacting with people by allowing one to reason on others' understanding of
the world in terms of visual perception, spatial descriptions, affordances and
beliefs, etc.  In the last years these notions have been gradually
employed in HRI.~\cite{Breazeal2006} presents a learning
algorithm that takes into account information about a teacher's visual
perspective in order to learn a task. ~\cite{Johnson2005} apply visual
perspective taking for action recognition between two
robots.~\cite{Trafton2005} use both visual and spatial perspective taking to
find out the referent indicated by a human partner.

\paragraph{Planning for cooperation} HRI offers new challenges to planning
by adding a complementary constraint: the synthesis of plans that are not only
feasible but legible (in terms of their intention) and compliant with human
preferences. This constraint is often ill-defined, dynamic, context-dependent.

\paragraph{Limits of disambiguation at semantic level}

\subsection{Joint actions with humans}
\label{sec:soa}

The human presence brings new requirements for robot's abilities both
at the functional and at the deliberative levels~\cite{Klein2004}. The
topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
in presence of humans as well as perception of human
activities~\cite{Breazeal2001,Burger2008}. Also, when
interacting with humans, robots need to incorporate communication and
collaboration abilities. Several theories dealing with
collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasize that
collaborative tasks have specific requirements compared to individual
ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realize it, they must show their
commitment to the goal during execution, etc. Several robotic systems
have already been built based on these
theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
all have shown benefits of this approach. They have also shown how
difficult it is to manage turn-taking between communication partners
and to interleave task realization and communication in a generic
way. Finally, today only few
systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
account at all levels.

\subsection{Knowledge-oriented architecture}

Altogether, the components we have presented compose a
\emph{knowledge-oriented} architecture:

\begin{itemize}

    \item{Knowledge is explicitly stored in one central and consistent
    repository of facts, accessible for all modules.}

    \item{Knowledge is represented in a strict formalism (OWL statements) and
    with a clearly defined vocabulary (stated in the common-sense ontology).}

    \item{The first two points enable a loosely-coupled architecture where
    modules can be removed or replaced easily by other ones as long as
    they share the same semantics (modules are defined by the knowledge they
    produce),}

    \item{We adopt a \emph{symbolic}, reactive, event-driven approach to robot control.
    By managing events at the same level as the reasoner, we take full
    advantage of the inference abilities of {\sc Oro} to trigger events whose
    \texttt{true} conditions can be inferred.}

    \item{Finally, this architecture allows for the combination of very
    different knowledge modalities in a single homogeneous environment,
    bringing mutual benefits to components. For instance, the dialogue
    processing module can perfectly run without any geometric
    perception, but its disambiguation routines can transparently
    benefit from it when available (since richer symbolic descriptions of
    objects are then available).}

\end{itemize}

Those items are not new \emph{per se}. It is however interesting to
underline the shift of focus this implies during the design and integration
phases of robots: components of our deliberative layer are defined and glued together by
the knowledge they produce and consume.

It reveals how the robotic community is ``catching up'' with semantic
technologies. Human-robot interaction, because it supposes operations in
environments with complex semantics, drives this evolution.

\subsection{Embodied Cognition}

@@

@@

We have presented in this article a decisional framework for human-robot
interactive task achievement that is aimed to allow the robot not only to
accomplish its tasks but also to produce behaviours that support its engagement
vis-a-vis its human partner and to interpret human behaviours and intentions.
Together and in coherence with this framework, we have developed and
experimented various task planners and interaction schemes that allow the robot
to select and perform its tasks while taking into account explicitly the human
abilities as well as the constraints imposed by the presence of humans, their
needs and preferences.

@@

There is also extensive work to be done in order to refine the notion of ``good
shared plan" and ``good/acceptable robot behaviour" in this context. There are
obviously large avenues for learning and adaptation in this context.

>>>>

\caption{The upper part of the {\sc Oro} common-sense TBox. All these concepts
    belong to the {\sc OpenCyc} namespace.}

@@

\paragraph{The OpenRobots Ontology} The {\sc Oro} common-sense ontology has been
designed from two requirements: covering our experimental needs and conforming
as much as possible to the {\sc OpenCyc} upper ontology.

@@

Figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
in the {\sc Oro} model between \emph{temporal} and \emph{spatial} entities (formally,
$(\text{TemporalThing} \sqcap \text{SpatialThing})^{\mathcal{I}} = \emptyset$, with
$\mathcal{I}$ the \emph{interpretation} of our model).

@@

The tree in Figure~\ref{fig|upper_tbox}\footnote{This subset of the ontology is
indeed a tree. This has however not to be the case in general, and, as a matter
of fact, the TBox of the whole {\sc Oro} common-sense ontology does not form a tree}
is not equally developed in each directions at lower levels. The descendants of
\concept{PartiallyTangibleThing} (\ie what we commonly call \emph{objects}), for
instance, are pictured in Figure~\ref{fig|tangible_things_tbox}. This excerpt
from the ontology makes the bottom-up design process clear: only few types of
\emph{partially tangible} things appear, and those are only subclasses relevant
to the context of service robotics in an human-like environment. For performance
reasons as well as clarity, we do not attempt to proactively extend the
conceptual coverage of the field. We instead opportunistically extend the
common-sense knowledge when required by experiments.

@@

Lastly, the {\sc Oro} common-sense ontology contains several rules and class
expressions that encode non-trivial inferences. The definition of
\concept{Bottle} as found in the {\sc Oro} ontology is a typical example:

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We call \emph{physical situation assessment} the
cognitive skill that a robot exhibits when it assesses the nature and content of its
surroundings and efficiently monitors its evolution.

%Spatial reasoning~\cite{O'Keefe1999} is a field in its own right, and has been
%used for natural language processing for applications such as direction
%recognition ~\cite{Kollar2010,Matuszek2010} or language
%grounding~\cite{Tellex2010}.~\cite{Skubic2004} present a spatial reasoner
%integrated in a robot which computes symbolic positions of objects.

We rely on a dedicated geometric and temporal reasoning module called {\sc
Spark} (\emph{SPAtial Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a
situation assessment reasoner that generates symbolic knowledge from the
geometry of the environment with respect to relations between objects,
robots and humans (Figures~\ref{fig|spark-oro}
and~\ref{fig:sparkSubfigures}), also taking into account the different
perspective that each agent has on the environment.

@@

%Our approach, implemented in the {\sc Spark} module, is a grounded amodal
%representation of the robot environment, where different sensing modalities
%(vision-based fiducial markers tracking, RGBD human tracking, motion capture)
%are merged and where amodal spatial reasoning takes place (for instance,
%explanation generation for unexpected perceptions like a disappearing object).

@@

{\sc Spark} is designed to run and continously update the knowledge base (in our
experiments, it runs at about 10Hz). At each steps, it re-computes spatial
relationships and affordances for the whole scene, and send the delta (new
relations and relations that do not hold anymore) to the knowledge base. This
approach may raise scalability concerns (we did not however hit such issues in
our constrainted scenarios), but greatly simplifies the management of the
\emph{dynamics} of the knowledge (\emph{When do I discard outdated knowledge?
When do I update it?}). Since it is equivalent to a reset of the reasoner
domain, it also hides issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).

@@

\paragraph{Symbolic Facts Production}

@@

\paragraph{Hypotheses on Objects States and Positions}

@@

\paragraph{Primitive Action Recognition}

@@

\subsubsection{Ego-centric and Allo-centric Frames}

@@

\subsubsection{Specific Limitations}

One of the main limit of our current approach is the way semantics are
conveyed: because we use 2D markers for artifacts, we skip most of the object
extraction and recognition challenges. Each object has a unique identifier,
which enable us to load a proper CAD model suitable for 3D spatial reasoning
and prevent recognition mistakes in the knowledge base. While {\sc Spark}
algorithms themselves are ignorant of the input sources, and would work equally
well with a full object recognition stack, we do not have tackled it yet.

@@

While parts of the architecture ({\sc Spark}, {\sc Oro}) have been deployed with external
execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
developed dedicated robot controllers which integrate the whole stack introduced
in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
written in the PRS~\cite{Ingrand1996} language. We also use the Python-based
{\sc pyRobots} collection of tools that provide a large set of high-level
actions and an event-based architecture well suited for prototyping.
They both rely on deep integration with the knowledge base that serves as the
primary source of semantics for decision making.

The novelty of our contribution here is the interleaving of a semantic-aware
architecture designed for autonomy with human-aware decisional components. To
our knowledge, no other robotic architecture covers in depth those two
perspectives.

The Ke Jia project~\cite{Chen2010} is maybe one of the closest
approach. It integrates on a mobile platform a knowledge representation language
with natural language processing, task planning and motion planning. Default and
non-monotonic reasoning has been especially researched within the Ke Jia project
for symbolic task planning~\cite{Ji2011} and underspecified natural language
processing. The Ke Jia robot has been demonstrated in several tasks involving human-robot
interaction with natural language. These tasks include a task with multiple
\emph{pick \& carry} that are globally optimised, naive physics reasoning via
taught rules or more complex scenarios with the robot delivering drinks, taking
into account changing and mutually exclusive preferences of users.

Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
called \textsc{CRAM} (Cognitive Robot Abstract Machine) that deeply integrates
with \textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework
based on Prolog. Its underlying storage is based on an OWL ontology, derived
from \textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated
on several real-world scenarios, where natural language recipes extracted from
the Internet had to be translated into plans and executed in a kitchen
environment, perceived and rebuilt on-line by the robots. This architecture has
however not been deployed in many scenarios involving human-robot interactions.

@@

The second one involves the {\sc Shary} execution controller and the HATP symbolic
task planner besides {\sc Oro} and {\sc Dialogs}. In this scenario, the human and the
robot try to cooperatively remove objects from a table. behaviours and motions
are fully planned.

@@

Perception of objects is done through a tag-based system and humans are
detected through motion capture. The robot knowledge base is pre-loaded with
the \emph{{\sc Oro} Commonsense Ontology}.  We next describe two
situations where we can follow the internal robot's reasoning and the
interaction with the users.

@@

This experiment involved a more complex decisional layer where the {\sc Oro} server
was used in conjunction with the HATP symbolic task planner and the {\sc Shary}
execution controller (the complete software architecture we deployed for this
experiment is the one pictured in Figure~\ref{fig|archi},
page~\pageref{fig|archi}).

@@

\section{Discussion: when Artificial Intelligence enables Human-Robot
Interaction}
\label{sect|conclusion}

This last section summarises our main contributions and tries to clarify the ``take home'' message: what are the challenges
that human-robot interaction brings to artificial intelligence, and what is the
current state-of-the-art in this field?

@@

\subsubsection{Embodied Cognition}

@@

\subsection{Knowledge and Robotics}
\label{krs-discussion}

\subsubsection{RDF as a Formalism for Semantics}

Knowledge flows in our architecture through a \emph{semantic blackboard}, the
{\sc Oro} server. We rely on Description Logics (OWL) to represent and
manipulate facts.

Alternatives like \emph{Answer Set Programming} have also been
successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
to deal with non-monotonic reasoning. However we did not actually hit any brick wall
while working with OWL ontologies. We may reconsider this choice at a later
stage, but until now it has proven an effective framework to quickly explore
implementations of new cognitive abilities (for instance, it has been
conceptually and technically easy to add support for independent knowledge
models, one per agent the robot interacts with --see section~\ref{sect|tom}).

\subsubsection{Knowledge and Architecture}

As a whole, the components that we have introduced so far build a
knowledge-oriented architecture: Knowledge is explicitly stored in one central
and consistent repository of facts, accessible for all modules. It relies on a
\emph{strict formalism} (OWL statements), with a \emph{well defined vocabulary}
(stated in the common-sense ontology). These first two points lead to a
\emph{loosely-coupled architecture} where modules can be removed or replaced
easily by other ones as long as they share the same semantics (modules are
defined by the knowledge they produce).

Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
managing events at the same level as the reasoner, we take full advantage of the
inference abilities of {\sc Oro} to trigger events whose \texttt{true}
conditions can be inferred.

And finally, this architecture allows for the combination of very different
knowledge modalities in a single homogeneous environment, bringing mutual
benefits to components. For instance, the dialogue processing module can
perfectly run without any geometric perception, but its disambiguation routines
can transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).

Those items are not new \emph{per se}. It is however interesting to underline
the shift of focus this implies during the design and integration phases of
robots: components of our deliberative layer are defined and glued together by
the knowledge they produce and consume. Human-robot interaction, because it
supposes operations at \emph{human-level} and in environments with complex
semantics, acts here as a motivational force.

\subsection{Bringing the Humans into the Equation}

\subsubsection{Modelling the Humans}

Perspective taking is this human ability which allows one to put him/herself in
another person's point of view. Studied in psychology
literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
interacting with people by allowing one to reason on others' understanding of
the world in terms of visual perception, spatial descriptions, affordances and
beliefs, etc.  In the last years these notions have been gradually employed in
HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
information about a teacher's visual perspective in order to learn a task.
~\cite{Johnson2005} apply visual perspective taking for action recognition
between two robots.~\cite{Trafton2005} use both visual and spatial perspective
taking to find out the referent indicated by a human partner.

\subsubsection{Acting in Presence of Humans}

HRI offers new challenges to planning
by adding a complementary constraint: the synthesis of plans that are not only
feasible but legible (in terms of their intention) and compliant with human
preferences. This constraint is often ill-defined, dynamic, context-dependent.

\paragraph{Joint actions with humans}\label{sec:soa}
The human presence brings new requirements for robot's abilities both
at the functional and at the deliberative levels~\cite{Klein2004}. The
topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
in presence of humans as well as perception of human
activities~\cite{Breazeal2001,Burger2008}. Also, when
interacting with humans, robots need to incorporate communication and
collaboration abilities. Several theories dealing with
collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasize that
collaborative tasks have specific requirements compared to individual
ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realize it, they must show their
commitment to the goal during execution, etc. Several robotic systems
have already been built based on these
theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
all have shown benefits of this approach. They have also shown how
difficult it is to manage turn-taking between communication partners
and to interleave task realization and communication in a generic
way. Finally, today only few
systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
account at all levels.

\paragraph{Limits of Disambiguation at Semantic Level}

@@

%We have presented in this article a decisional framework for human-robot
%interactive task achievement that is aimed to allow the robot not only to
%accomplish its tasks but also to produce behaviours that support its engagement
%vis-a-vis its human partner and to interpret human behaviours and intentions.
%Together and in coherence with this framework, we have developed and
%experimented various task planners and interaction schemes that allow the robot
%to select and perform its tasks while taking into account explicitly the human
%abilities as well as the constraints imposed by the presence of humans, their
%needs and preferences.

@@

There is also extensive work to be done in order to refine the notion of ``good
shared plan'' and ``good/acceptable robot behaviour'' in this context. There are
obviously large avenues for learning and adaptation in this context.
